Almost orthogonal vectors

### 2014-12-11 project3/paper/paper.tex
more edits through part of Section 6

\section{Some Optimal Configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{$n=3,N=6$}
Let us consider collections of six vectors in $\mathbb R^3$. As can be seen in Table 1, the lower bound provided in this case by Theorem \ref{lower_bound} is $\frac 1 5$ as well. We shall now see, in fact, that this lower bound is attainable.
\begin{theorem}
For $n=3$ and $N=6$, the optimum value of $\epsilon$ is given by $\frac 1 5$.
\end{theorem}
\begin{proof}Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$. This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$. It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$. If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$. Coupled with our knowledge that $\frac 1 5$ is a lower bound on $\epsilon$, we conclude that these six points give us an optimal configuration.\end{proof}
\subsection{$N=n+1$}
Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so it suffices to show there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$.

@@

Since visualizing in spaces of dimensionality $3$ or greater is extremely hard, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs. We describe this method in greater detail below.

@@

Even though this approach seems simple, it actually produces remarkably precise values -- this can be verified by cross-checking with results we've already proven / conjectured in this paper, this can be seen from values in Table $2$.

In particular, the computed values for $n=2$ are pretty accurate -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.

@@

\section{Variations}

We now consider a variation of the above problem, where we actually try to place $N$ vectors in $n$-space as far apart as possible from each other. More concretely, if we define $\delta$ as

$$\delta = \max_{i \neq j} v_i \cdot v_j$$,

then we can formalize our problem as finding the $N$ vectors $v_1, v_2, \ldots, v_N$ that minimize $\delta$.

Before proving what the optimum value of $\delta$ is for $N= n+1$, we introduce the parallelogram law.

>>>>

\section{Some optimal configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{$n=3,N=6$}
Let us consider collections of six vectors in $\mathbb R^3$. As can be seen in Table 1, the lower bound provided in this case by Theorem \ref{lower_bound} is $\frac 1 5$ as well. We shall now see, in fact, that this lower bound is attainable.
\begin{theorem}
For $n=3$ and $N=6$, the optimum value of $\epsilon$ is given by $\frac 1 5$.
\end{theorem}
\begin{proof}Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$. This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$. It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$. If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$. Coupled with our knowledge that $\frac 1 5$ is a lower bound on $\epsilon$, we conclude that these six points give us an optimal configuration.\end{proof}
\subsection{$N=n+1$}
Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so it suffices to show there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$.

@@

Because of the difficulty involved with visualization in dimensions 3 and above, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs. We describe this method in greater detail below.

@@

Though this approach seems simple, it produces  results which seem to agree with the $\epsilon$ we derived for $(2, N)$ in Section 2, and with the $\epsilon$ we derived for $(3, 6)$ and $(n, n+1)$ in Section 4. This gives us some confidence in the accuracy of our algorithm. However, it must be noted that the data do not agree with the bounds produced by Theorem \ref{lower_bound}. There are multiple interpretations of this fact. First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight; they in fact are not for the case $(2, N)$. Alternatively, it is possible that our code simply becomes inaccurate for larger $n$ and $N$. This could be a fruitful direction for further research.

@@

\section{A variation of the problem}

We now consider a variation of the above problem, where we actually try to place $N$ unit vectors in $n$-space as far apart as possible from each other. More concretely, we define $\delta$ as
$$\delta = \max_{i \neq j} v_i \cdot v_j,$$
and we formalize our problem as finding the $N$ vectors $v_1, v_2, \ldots, v_N$ that minimize $\delta$.

Our results on this section concern the case when $N=n+1$. We shall see, in fact, that the optimal value of $\delta$ is given by $-\frac{1}{n}$, we introduce the parallelogram law.
